Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-09 paper.tex

\begin{figure} \centering
\includegraphics[width=0.4\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra}
Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}. Mutations are
categorised by type (i.e., ancestral state $>$ derived state). The percentages
of each mutation type from the wide ARG are represented by blue bars and the
percentages from Yi et al. by orange bars, with the darker colours representing
one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
\end{figure}

@@

\subsubsection{Complex origins} 
\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node. The samples descending from these nested
recombinations include 9 that are assigned by Nextclade to
various non-recombinant pre-Alpha lineages (blue).}
\end{figure}

As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which multiple non-independent (``nested'') recombination events are involved.
In these cases, the descendants of the original recombination
frequently include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are mostly not
identified as recombinants by Nextclade, and plotted in blue. A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Investigating examples of complex recombinant origins, and identifying which
(if any) of the nested recombination events may be artefactual, is an important
area of future research.

>>>>

\begin{figure} \centering
\newcolumntype{Y}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{Yc}

\begin{tabular}[b]{llrlr}\toprule
            & \multicolumn{2}{l}{Wide ARG} & \multicolumn{2}{l}{Long ARG} \\
% \midrule
    \cmidrule{2-5}
Total      & \multicolumn{2}{l}{1,213,193} & \multicolumn{2}{l}{1,062,072} \\
Transitions & 873,487   & (72.00\%) & 783,773  & (73.80\%) \\
Transversions & 326,053 & (26.88\%) & 270,333  & (25.45\%) \\
Insertions  & 6191  & (0.51\%) & 2814   & (0.26\%) \\
Deletions   & 7462  & (0.62\%) & 5152   & (0.49\%) \\
Recurrent   & 74,719 & (6.16\%) & 50,099 & (4.72\%) \\
Reversions  & 72,617 & (5.99\%) & 48,226 & (4.54\%) \\
Private     & 758,903 & (62.55\%) &  767,111 & (72.23\%)\\
\bottomrule
\end{tabular}
&
\includegraphics[width=0.35\textwidth]{figures/mutational_spectra.pdf}\\
(A) & (B)\\
\end{tabularx}
\caption{\label{fig:mutational_spectra}
(A) Summary of mutations in the Long and Wide ARGs.
(B) Mutational spectrum of SARS-CoV-2 in
the wide ARG compared to \cite{Yi2021-sc}. Mutations are
categorised by type (i.e., ancestral state $>$ derived state). The percentages
of each mutation type from the wide ARG are represented by blue bars and the
percentages from Yi et al. by orange bars, with the darker colours representing
one direction (e.g., C$>$U) and the lighter colours the reverse (e.g., U$>$C).}
\end{figure}

@@

\subsubsection{Complex origins}
As well as cases where single recombination events explain an X lineage,
such as the examples in Figure~\ref{fig:pango-simple-origin-graph}, many of the
Pango X lineages involve more complex origins,
in which multiple non-independent (``nested'') recombination events are involved.
In these cases, the descendants of the original recombination
frequently include samples which have not been previously identified as recombinants.
An example is XB in the long ARG,
plotted in Figure~\ref{fig:complex_origins_graph}. Here, a recombination
between a B.1 sample and B.1.627 sample leads both to the XB samples and to a
``hairball'' of further recombination nodes whose descendants are mostly not
identified as recombinants by Nextclade, and plotted in blue. A similar
pattern is seen when examining XB in the wide ARG (discussion above, plot not shown).
Note that the descendants of nested recombination events include only one
XB sample (pink upper left, with 7 mutations above it); moreover, this sample is
not identified as XB by GISAID (Supplementary Figure~\ref{fig:pango_XB_gisaid_graph})
indicating some uncertainty in lineage assignment in this part of the ARG.
Investigating examples of complex recombinant origins, and identifying which
(if any) of the nested recombination events may be artefactual, is an important
area of future research.

\begin{figure} \centering
\includegraphics[width=\textwidth]{figures/Pango_XB_nxcld_tight_graph.pdf}
\caption{\label{fig:complex_origins_graph}  A Pango lineage subgraph showing
complex origins of the recombinant lineage XB. All XB samples in the long ARG
trace to a single recombination node (top centre), but three further recombinations
also descend from this node. The samples descending from these nested
recombinations include 9 that are assigned by Nextclade to
various non-recombinant pre-Alpha lineages (blue).}
\end{figure}
